A Computational Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2017-03-29 paper/paper.tex
superpass

\author{Elvis Dohmatob, Bert Kappen, Danilo Bzdok\\
  %% INRIA, Parietal team, Saclay, France\\
  %% CEA, Neurospin, Gif-sur-Yvette, France\\
  %% firstname.lastname@inria.fr
}

@@

\begin{abstract}
The default mode network (DMN) is the brain manifestations of the human
baseline mental activity.
%
Many research streams agree on its likely role in evolutionarily adaptive
envisioning of mental scenarios to predict the environment.
The DMN would hence be dedicated to continuous autobiographical memory retrieval,
generation of hypothetical outcomes,
and reward contingency evaluation when letting the mind go.
It explains its highest energy consumption in the brain and
its intimate coupling with conscious awareness.
%
This concept paper proposes a process model that describes
\textit{how} the DMN may actually implement continuous
environmental assessment and prediction to guide action choices.
DMN function is recast in mathematical terms
from the perspective of
% control theory and
reinforcement learning.
We argue that our formal account of DMN function
explains existing cognitive accounts as special cases and
parsimoneously explains a variety of experimental findings
in animals and humans.
%
Formal models of the computational processes subserved in the DMN
could offer access routes to the statistical regularities
of human predictive behavior.
\end{abstract}

% official NIPS keywords
\textbf{\\keywords}: systems biology, mind wandering, cognitive science,
artificial intelligence (AI),
reinforcement learning
% autonomous learning systems
% \textbf{\\Get opinion from}: Danielle Bassett, Kai Vogeley, Mohammad Shakir,
% Guillaume Dumas, Smallwood, Tenenbaum

@@

\section{Introduction}
%
When left unperturbed, the human brain is not at rest.
Rather, the brain continues to metabolize large quantities of
oxygen and glucose energy to maintain
potentially purposeful neuronal computation
in the absence of a focused behavioral goal
\citep{kenet2003spontaneously, fiser2004small}.
This baseline energy demand is subject to surprisingly little modulations
when mentally processing environmental challenges
\citep{raichle2001pnas}.
What has early been described as the "stream of consciousness"
in psychology \citep{james1890principles}
found its neurobiological manifestation
in the so-called "default mode network" (DMN)
\citep{shul1997}. This networked collection of some of the highest regions in
the human association cortex \citep{mesulam1998sensation, margulies2016situating}
consistently
increased in neural activity during
unfocused everyday mind wandering \citep{raichle2001pnas}.
In the beginning of the 21st century, brain imaging was hence the first technology
to allow for the discovery of a unique brain network that subserves
baseline mental activities
\citep{bzdok2015resting}.

@@

\subsection{Sketch of our main contributions}
Our theory proposes that the implication of the DMN in a diversity of human cognitive processes
can be parsimoniously recast as prediction error minimization
based on pervasive probabilistic simulations,
thus maximizing reward outcome at various time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo sampling
\citep{tenenbaum2011grow}.
\textit{Control} refers to the influences that an agent exerts when interacting
with the environment to encourage preferred states.
Even (necessarily imperfect) memories
of past experience, random mind-wandering and dreams during sleep
may provide meaningful building blocks to iteratively improve
the predictive DMN machinery to optimize the behavior of the organism.
%
In short, it has been proposed earlier that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
The idea has been invigorated by empirical evidence from
neuroscientific experiments \citep{kording2004bayesian, fiser2004small}.
This paper proposes a formal model that satisfies this contention.

@@

\section{Reinforcement Learning, a computational model for DMN function}
We now argue that the DMN is equipped with sufficient neurobiological substrates gradient
for implementing all the major components (memory, reward signaling, accumulators, random sampling, etc.)
of an efficient \textit{reinforcement learning (RL)} algorithm, and all these
components integrate in a plausible way.
Reinforcement learning is learning to achieve a goal by interacting with an external
\textit{environment}. In a general RL framework, an \textit{agent}
interacts with an  environment in a trial-by-error fashion by taking
actions $a$, which trigger changes in the
\textit{state} of the environment
$s \rightarrow s'$, accompanied by a real-valued \textit{reward}
$r = r(s, a)$ (or regret!, since it can be negative) collected by the
agent. In this view, the environmental is partially controlled by
the action of the agent and the reward should be thought
of as an instantaneous satisfaction accompanying the execution of
an action, and may be delayed, only becoming substantial when accumulated over an extended period of time.
%% Here the subscript $t$ denotes time, and can be discrete or
%% continuous.
The environment is generally taken as stochastic,
is \textit{partially observed},
in the sense that only part of the current state is observed by
the agent. This unpredictability
% (in a first-order sense)
is realistic in a model which sets out
to explain DMN function. Indeed, it is known that, the more the external world is predictable,
the more mental activity becomes detached from the actual sensorium
\citep{antrobus1966studies, pope1978regulation}.
Conversely, the more the currently executed task is unknown and unpracticed,
the lower the DMN activity, i.e the tendency for stimulus-independent thoughts
\citep{filler1973daydreaming, teasdale1995stimulus}.
%
These ``offline'' processes may however contribute to optimized control of the organism.
A DMN framework incorporating reinforcement learning
can naturally embed human behavior
into the tension between exploitative action with immediate gains and
explorative action with longer-term reward schedules
\citep{dayan2008decision}.
The implication of the DMN in a diversity of human cognitive processes
can be parsimoneously recast as prediction error minimization
based on pervasive probabilistic simulations,
thus maximizing reward outcome at various time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo sampling
\citep{tenenbaum2011grow}.

\subsection{Partially Observable Markov Decision Processes}
It is our view that at a sufficiently coarse scale, the brain is a physical system governed by the laws of
\textbf{XXX the paper is about the DMN more specifically, perhaps try to sharpen}
physics and can be locally described by Markov processes. Indeed, as noted in \cite{tegmark2016improved},
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short, defining $s(t)$, the state at time $t$, as the position in phase space.
If the process has $memory$ such that the next state depends not only on the current state but also on some finite
number of past states,
rational probabilistic planning can reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include elements of the past.

In artificial intelligence (AI), a popular abstract model for
multi-step decision processes in such an environment are so-called
\textit{Partially Observable Markov Decision Processes (POMDPs)}.
A POMDP models a decision process in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state. Instead, it tries to optimize a \textit{subjective} reward
signal which is tried alongside each transition in the environments state-space, and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities. This is a minimal amount of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL}, the approach adapted by our work.
\begin{mdframed}
\textbf{Model-free RL in the brain is plausible.}  Indeed, as was noted in \cite{dayan2008decision}, it has long been suggested in the literature that there
  is a rather direct mapping of model-free RL learning algorithms
onto the brain, with the neuromodulator dopamine serving
as a teaching signal to train values or policies by controlling
synaptic plasticity at targets such as the ventral and
dorsolateral striatum.
\end{mdframed}

In contrast, \textit{model-based RL} starts off with some mechanistic assumptions about the dynamics of world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state-space and transitions thereupon.
For example if a rat in a maze knows, for sure,
that staying still will produce no change in the environment, and in particular will not eventually lead to finding the food.
% A core property of human intelligence has been proposed underly to improve
% on an expected utility outcomes as a strategy for action choice in uncertain
% environments \citep{gershman2015computational}.
Our rat might represent such knowledge about the world as:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a cell / chamber containing food.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}

Mathematically, a POMDP is simply a quintuplet $(\mathcal S, \mathcal A, r, p, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, e.g $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $p : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto p(s'|s,a)$,
  the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such
  transitions be Markovian. Consequently, the future states are independent of past states and only depend on the present.
\item $\mathcal A$ is the set of actions, e.g $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$ is the \textit{reward function},
  so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts-off in state $s$.
\end{itemize}

@@

\subsection{Long-term rewards and policies}
The behavior of the agent is governed by some kind of \textit{policy},
which maps states of the world to
a set of candidate actions to perform given this state. Starting a time $t=0$,
a policy $\pi$ generates a trajectory as follows
\begin{eqnarray*}
  \begin{split}
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect rewards }R_0 = r(s_0, a_0)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect rewards }R_1 = r(s_1, a_1)\\
    \ldots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect rewards }R_{t} = r(s_{t}, a_{t})\\
    \ldots
  \end{split}
\end{eqnarray*}
Since an action taken now might have repercussions long into the feature, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
cumulative reward estimate which takes into account the future. A standard for
modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t.
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following a policy $\pi$.
\begin{mdframed}
  \textbf{Where is ``value'' buffered in the brain ? You mean where in the DMN!
  then the answer is in the vmPFC which is connected to the NAc}
  The amygdala is known to be involved in significance evaluation) and
  the nucleus accumbens (NAc) is involved in reward evaluation.
\end{mdframed}
  The  goal of RL is then adapt this policy so as to  maximize the cumulative rewards. In
  the definition of $G^\pi$ above, the constant $\gamma$ $(0 \le \gamma < 1$) is the reward \textit{discount factor}.
Taking $\gamma = 0$   corresponds to a perfectly myopic agent who is solely concerned about their
immediate rewards and has no horizon for the future, which is not compatible with long-term goal planning
as potentially implemented in the DMN. For there to be any learning at all, it is thus necessary that $0 < \gamma < 1$.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus, the agent is positively biased towards considering rewards which with
a short time delay. The reward importance hence decays exponentially
with distance in the future,
given that there is more uncertainty in the farsighted future.
The reward discount factor $\gamma$ can thus be seen as calibrating
the delay discounting behavior of the intelligent agent,
that is the decision related to reward-delay tradeoff schedules.
It is also important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
For instance, a deterministic strategy in playing rock-paper-scissors can
be easily exploited and a uniformly random choice is optimal.
\textbf{XXX I love this example :-)}

\subsection{The Q-value: a goodness measure for actions}
The agent's goal can then be formulated as finding policy $\pi$ which maximizes the
expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a,a_1 \sim \pi,a_2 \sim \pi,\ldots].
\end{equation}
In plain English, the Q-value $Q^{\pi}(s,a)$ is the expected / average / mean future reward
over all possible trajectories in which the agent starts off starts at with an environment in state
$s$, takes and action $a$, and then follows the policy $\pi$ thereafter to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action; it answers the question
"What is the expected utiltiy of taking action $a$ in this situation ?".
When a choice of actions is available, the $Q^{\pi}(s,a)$
provides a formalization of optimal selection.

\subsection{Computing optimal behavior}
By construction, maximizing the expectation in  \eqref{eq:qval} is computationally intractable in general
due to the enormous number of possible action
trajectories the agent might need to
consider.
% \begin{mdframed}
%   \textbf{N.B.:}
%   If the Bellman-based computational ideas have a corresponding
%   neurobiological implementation, this provides a framework of how
%   optimal behavior could be computable in the human brain.
% \end{mdframed}
One popular simplification that removes
this issue is to only consider the sub-family of
\textit{deterministic policies}, which map each state to the best single best action which
should be taken at thin state, and therefore define functions from states to actions.

\subsubsection{Q-learning} Q-learning (Watkins \& Dayan, 1992), a commonly used off-policy algorithm, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The \textit{Bellman's equation} \cite{sutton1998reinforcement} then takes the simple form
\begin{equation}
  \begin{split}
    Q^\pi(s, a) &= \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q^\pi(s', a')]
    \\
    &= \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma Q^\pi(s', \pi(s'))],
  \end{split}
  \label{eq:bellman}
\end{equation}
a \textit{fixed-point} equation which provides a recursive decomposition of optimal behavior.
A complicated dynamic programming optimization can be
broken into simpler sub-problems at different time points,
which has often been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value,
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Notice that in the expectation in \eqref{eq:bellman}, the sampling is done over only things which
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

\subsubsection{Efficient learning via value approximation and experience replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
A successful RL theory needs ways to approximate the state-action value function
which scale up to large state spaces and / or action spaces.
In state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new state $s' \in \mathcal S$. At the end of
this step, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar
behavior of the agent and is recorded in replay memory buffer:
$D \leftarrow D \cup \{e\}$ (possibly discarding the oldest entries to make space).
Now, at iteration $k+1$, replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$ (hyppocampus ?)
and trying to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q_k)$, with the output of a parametrized regression model $(s,a,r,s')
\mapsto {Q}(s, a|\theta^Q_{k})$, where $Q(.,.|\theta^Q_{k})$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q_{k+1}$.

In the brain, such an experience replay mechanism is probably implemented in the hyppocampus (HC). Indeed,
hippocampal replay is a phenomenon observed in rats, mice, cats, rabbits, songbirds, and monkeys. Indeed, during sleep or awake rest, replay refers to the re-occurrence of a sequence of cell activations that also occurred during activity, but the replay has a much faster time scale.

>>>>

\author{Elvis Dohmatob, Guillaume Dumas, Danilo Bzdok\\
  %% INRIA, Parietal team, Saclay, France\\
  %% CEA, Neurospin, Gif-sur-Yvette, France\\
  %% firstname.lastname@inria.fr
}

@@

\begin{abstract}
The default mode network (DMN) is the brain manifestation of the human
baseline mental activity.
%
Many research streams agree on its likely role in evolutionarily adaptive
envisioning of mental scenarios to predict the environment.
The DMN would hence be dedicated to continuous autobiographical memory retrieval,
generation of hypothetical outcomes,
and reward contingency evaluation when letting the mind go.
It explains its highest energy consumption in the brain and
its intimate coupling with conscious awareness.
%
This concept paper proposes a process model that describes
\textit{how} the DMN may actually implement continuous
environmental assessment and prediction to guide action choices.
DMN function is recast in mathematical terms
from the perspective of
% control theory and
reinforcement learning.
We argue that our formal account of DMN function
explains existing cognitive accounts as special cases and
parsimoneously explains a variety of experimental findings
in animals and humans.
%
Formal models of the computational processes subserved in the DMN
could offer access routes to the statistical regularities
of human predictive behavior.
\end{abstract}

% official NIPS keywords
\textbf{\\keywords}: systems biology, mind wandering, cognitive science,
artificial intelligence,
reinforcement learning
% autonomous learning systems
% \textbf{\\Get opinion from}: Danielle Bassett, Kai Vogeley, Mohammad Shakir,
% Guillaume Dumas, Smallwood, Tenenbaum

@@

\section{Introduction}
%
When left unperturbed, the human brain is not at rest.
Rather, the brain continues to metabolize large quantities of
oxygen and glucose energy to maintain
potentially purposeful neuronal computation
in the absence of a focused behavioral goal
\citep{kenet2003spontaneously, fiser2004small}.
This baseline energy demand is subject to surprisingly little modulations
when mentally processing environmental challenges
\citep{raichle2001pnas}.
What has early been described as the "stream of consciousness"
in psychology \citep{james1890principles}
found its neurobiological manifestation
in the so-called "default mode network" (DMN)
\citep{shul1997}. This networked collection of some of the highest regions in
the human association cortex \citep{mesulam1998sensation, margulies2016situating}
consistently
increased in neural activity during
unfocused everyday mind wandering \citep{raichle2001pnas}.
Indeed, the human brain features "a hierarchy of brain systems with
the DMN at the top and the salience and dorsal attention systems
at intermediate levels, above thalamic and unimodal sensory
cortex" \citep{carhart2010default}.
In the beginning of the 21st century, brain imaging was hence the first technology
to allow for the discovery of a unique brain network that subserves
baseline mental activities
\citep{bzdok2015resting}.

@@

\subsection{Sketch of our main contributions}
Our theory proposes that the implication of the DMN in a
diversity of humans' most advanced cognitive processes
can be parsimoniously recast as prediction error minimization
based on pervasive probabilistic simulations,
thus maximizing reward outcome at various time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo sampling
\citep{tenenbaum2011grow}.
\textit{Control} refers to the influences that an agent exerts when interacting
with the environment to encourage preferred states.
Even (necessarily imperfect) memories
of past experience, random mind-wandering and dreams during sleep
may provide meaningful building blocks to iteratively improve
the predictive DMN machinery to optimize the behavior of the organism.
%
In short, it has been proposed earlier that
the human brain's energy budget is largely dedicated to
``the development and maintenance of [a]
probabilistic model of anticipated events''
\citep{raichle2005intrinsic}.
The idea has been invigorated by empirical evidence from
neuroscientific experiments \citep{kording2004bayesian, fiser2004small}.
This paper proposes a formal model that satisfies this contention.

@@

\section{Reinforcement learning: A process model for DMN function}
The reviewed neurobiological properties and likely functions of the DMN
are now argued to be sufficient
for implementing all the major components 
of a full-fledged \textit{reinforcement learning (RL)} algorithm.
Memory, reward signaling,
information gathering, random sampling, and other components
of intelligent RL agents appear to cross-talk in the DMN in a plausible way.

Reinforcement learning is learning to achieve a goal by interacting with an external
\textit{environment}.
\textbf{XXX Don't define something with the same thing}
In a general RL framework, an \textit{agent}
interacts with an  environment in a trial-by-error fashion by taking
\textit{actions} $a$, which trigger changes in the
\textit{state} of the environment
$s \rightarrow s'$, accompanied by a real-valued \textit{reward}
$r = r(s, a)$ (or regret!, since it can be negative) collected by the
agent.
In this view, the environment is partially controlled by
the action of the agent and the reward should be thought
of as an instantaneous
\textbf{XXX instant is problematic here, because it may change behavior only later in RL}
satisfaction
\textbf{XXX or punishment?}
accompanying the execution of
an action, and may be delayed, only becoming substantial when accumulated over an extended period of time.
%% Here the subscript $t$ denotes time, and can be discrete or
%% continuous.
The environment is generally taken as stochastic,
\textbf{XXX One sentence, one bullet. Please explain stochast and PO
in one sentence each.}
is \textit{partially observed},
in the sense that only part of the current state is observed by
the agent.
This unpredictability
\textbf{XXX Explain this sentence better because it is important.}
is realistic in a model which sets out
to explain DMN function. Indeed, it is known that, the more the external world is predictable,
the more mental activity becomes detached from the actual sensory inputs
\citep{antrobus1966studies, pope1978regulation},
the more stimulus-independent thoughts occur, and
the higher the neural activity in the DMN.
Conversely, the more the currently executed task is unknown and unpracticed,
the lower the DMN activity
\citep{filler1973daydreaming, teasdale1995stimulus}.
%
These ``offline'' processes may however contribute to
optimizing control of the organism.
A DMN framework incorporating reinforcement learning
can naturally embed human behavior
into the tension between exploitative action with immediate gains and
explorative action with longer-term reward schedules
\citep{dayan2008decision}.
The implication of the DMN in a diversity of human cognitive processes
can be parsimoneously explained as prediction error minimization
coupled with probabilistic simulations in form of mental scenes,
thus calibrating action portfolios and
maximizing reward outcome at various time scales.
Such a purposeful optimization objective
may be solved by a stochastic approximation
based on a brain implementation of Markov Chain Monte Carlo (MCMC) sampling
\citep{tenenbaum2011grow}.

\subsection{Partially Observable Markov Decision Processes}
It is our view that at a sufficiently coarse scale, the brain is a physical system governed by the laws of
\textbf{XXX the paper is about the DMN more specifically, perhaps try to sharpen}
physics and can be locally described by Markov processes. Indeed, as noted in \cite{tegmark2016improved},
any system obeying the laws of classical physics can be accurately modeled as a Markov process as long as the time
step is sufficiently short, defining $s(t)$, the state at time $t$, as the position in phase space.
If the process has $memory$ such that the next state depends not only on the current state but also on some finite
number of past states,
rational probabilistic planning can be reformulated
as a standard memoryless Markov process by simply expanding the
definition of the state $s$ to include elements of the past.

In artificial intelligence (AI), a popular abstract model for
multi-step decision processes in such an environment are so-called
\textit{Partially Observable Markov Decision Processes (POMDPs)}.
A POMDP models a decision process in which it is assumed that environment dynamics are determined by a Markov process,
but the agent cannot directly observe the underlying state. Instead, it tries to optimize a \textit{subjective} reward
signal. This is performed alongside each transition in the environments state-space, and maintains a probability distribution over the set of possible
states, based on a set of observations and observation probabilities. This is a minimal amount of assumptions that can be made about an environment,
and is characteristic of so-called \textit{model-free RL}, the approach adapted by our work.
Such model-free RL is plausible in the human brain. Indeed, as was noted in \cite{dayan2008decision}, it has long been suggested in the literature that there
  is a rather direct mapping of model-free RL learning algorithms
onto the brain,
\textbf{XXX New sentence}
with the neuromodulator dopamine serving
as a teaching signal to train values or policies by controlling
synaptic plasticity at targets such as the ventral and
dorsolateral striatum.
\textbf{XXX Loose all boxes by integrating the text please since it is not a usual layouting}

In contrast, \textit{model-based RL} would start off with some mechanistic assumptions about the dynamics of world.
These can be assumptions about the physical laws governing the agent's world or constraints on the state space and transitions between states.
For instance, if a rat in a maze knows
that standing still will produce no change in the environment, and in particular will not eventually lead to finding the food.
% A core property of human intelligence has been proposed underly to improve
% on an expected utility outcomes as a strategy for action choice in uncertain
% environments \citep{gershman2015computational}.
\textbf{XXX transition back to model-free RL is not clear. please make explicit}
Our rat might represent such knowledge about the world as:
\begin{itemize}
\item $r(s, \text{``stand still''}) = 0$ if $s$ does not correspond to a cell / chamber containing food.
\item $p(s'|s,\text{``stand still''}) = 1$ if $s'=s$ and $0$ otherwise.
\item etc.
\end{itemize}

Mathematically, a POMDP is simply a quintuplet $(\mathcal S, \mathcal A, r, p, \mu)$ where
\begin{itemize}
\item $\mathcal S$ is the set of states, such as $\mathcal S = \{\text{happy}, \text{sad}\}$.
\item $p : \mathcal S \times \mathcal A \times \mathcal S \rightarrow [0, 1],\; (s,a,s') \mapsto p(s'|s,a)$,
  the probability of moving to state $s'$ if action $a$ is taken from state $s$. In addition, one requires that such
  transitions be Markovian. Consequently, the future states are independent of past states and only depend on the present.
\item $\mathcal A$ is the set of actions, such as $\mathcal A = \{\text{read}, \text{run},
  \text{laugh}, \text{sympathize}, \text{empathize}\}.$
\item $r : \mathcal S \times \mathcal A \rightarrow \mathbb R$ is the \textit{reward function},
  so that $r(s, a)$ is the instant reward for taking action $a$ in state $s$.
\item $\mu: \mathcal S \rightarrow [0, 1]$ is the prior probability on the states so that
  $\mu(s)$ is the probability that the environment starts off in state $s$.
\end{itemize}

@@

\subsection{Long-term rewards and action policies}
The behavior of the agent is governed by some kind of \textit{policy},
which maps states of the world to
a set of candidate actions to perform given this state. Starting a time $t=0$,
a policy $\pi$ generates a trajectory of action cascades as follows
\begin{eqnarray*}
  \begin{split}
    \textbf{ observe transition: }s_1 &\sim p(s|s_0,a_0)\textbf{ and collect rewards }R_0 = r(s_0, a_0)\\
    \textbf{choose action: }a_1 &\sim \pi(a|s_1)\\
    \textbf{ observe transition: }s_2 &\sim p(s|s_1,a_1), \textbf{ and collect rewards }R_1 = r(s_1, a_1)\\
    \ldots\\
    \textbf{choose action: }a_{t} &\sim \pi(a|s_{t})\\
    \textbf{ observe transition: }s_{t+1} &\sim p(s|s_{t},a_{t}), \textbf{ and collect rewards }R_{t} = r(s_{t}, a_{t})\\
    \ldots
  \end{split}
\end{eqnarray*}
Since an action taken in the present moment $t$
might have repercussions long into the future, it turns out that the
quantity to optimize is not the instantaneous rewards $r(s, a)$, but a
cumulative reward estimate which takes into account the future. A standard for
modeling this accumulation is the so-called
time-discounted \textit{cumulative reward} function
\begin{equation}
  \label{eq:cumr}
  G^\pi = \sum_{t=0}^{\infty}\gamma^{t}R_t.
\end{equation}
This random variable\footnote{Random as it depends both on the environment's dynamic and the
  policy $\pi$ being played (which can be stochastic).}  measures the cumulative reward of
following a policy $\pi$.
\begin{mdframed}
  \textbf{Where is ``value'' buffered in the brain ? You mean where in the DMN!
  then the answer is in the vmPFC which is connected to the NAc}
  The amygdala is known to be involved in significance evaluation) and
  the nucleus accumbens (NAc) is involved in reward evaluation.
\end{mdframed}
  The  goal of RL is then to adapt this policy so as to  maximize the cumulative rewards. In
  the definition of $G^\pi$ above, the constant $\gamma$ $(0 \le \gamma < 1$) is the reward \textit{discount factor}.
Taking $\gamma = 0$   corresponds to a perfectly myopic agent who is solely concerned about their
immediate rewards and has no horizon for the future, which is not compatible with long-term goal planning
as potentially implemented in the DMN.
To allow a learning process to arise,
it is thus necessary that $0 < \gamma < 1$.
% \footnote{In which case we can take $T \rightarrow +\infty$, provided the reward function $r$ is bounded.}.
Thus, the agent is positively biased towards considering rewards with
a short time delay. The reward importance hence decays exponentially
with distance in the future,
\textbf{XXX correct grammar in previous sentences}
given that there is more uncertainty in the farsighted future.
The reward discount factor $\gamma$ can thus be seen as calibrating
the delay discounting behavior of the intelligent agent,
that is the decision related to reward-delay tradeoff schedules.
It is also important to appreciate that a stochastic policy estimation
is more advantageous in many RL settings.
For instance, a deterministic strategy in playing rock-paper-scissors can
be easily exploited and a uniformly random choice is optimal.
\textbf{XXX I love this example :-)}

\subsection{The Q-value: A goodness measure for actions}
The purpose of the DMN can then be
formulated as finding a policy $\pi$ for the agent which maximizes the
expected cumulative value of an state-action pair $(a,s)$, also known as the \textit{Q-value}, namely
\begin{equation}
  \label{eq:qval}
  Q^{\pi}(s,a) = \mathbb E [G^\pi|s_0=s,a_0=a,a_1 \sim \pi,a_2 \sim \pi,\ldots].
\end{equation}
In other words,
the Q-value $Q^{\pi}(s,a)$ corresponds to the expected reward
over all possible action trajectories in which
the agent starts off starts at with an environment in state
$s$, chooses action $a$,
and then follows the policy $\pi$ thereafter to select future actions.
%% Given that each $\pi$ correspondence to an exhaustive description of
%% behavioral patterns, $Q^{\pi}(s_t,a_t)$ corresponds to the epistemic value of
%% the set of actions given states in finite time.
For the brain,
$Q^{\pi}(s, a)$ defined in \eqref{eq:qval} provides the subjective
utility of executing a specific action; it answers the question
"What is the expected utiltiy of taking action $a$ in this situation ?".
When a choice of actions is available, the $Q^{\pi}(s,a)$
provides a formalization of optimal behavior,
potentially driven by the DMN.

\subsection{Computing optimal behavior}
By construction, maximizing the expectation in  \eqref{eq:qval} is computationally intractable in general
due to the enormous number of possible action
trajectories the agent might need to
consider simultaneously.
% \begin{mdframed}
%   \textbf{N.B.:}
%   If the Bellman-based computational ideas have a corresponding
%   neurobiological implementation, this provides a framework of how
%   optimal behavior could be computable in the human brain.
% \end{mdframed}
One popular solution that remedies
this issue is to only consider the sub-family of
\textit{deterministic policies}, which map each state to the best single best action which
should be taken at thin state, and therefore define functions from states to actions.
\textbf{XXX sentence too long, please cut into 2}

\subsubsection{Q-learning} Q-learning (Watkins \& Dayan, 1992), a commonly used off-policy algorithm
\textbf{XXX WHATTTEEE?}
, uses the
\textit{greedy policy}
\begin{equation}
  \pi(s) = \argmax_{a \in \mathcal A}{Q}(s, a).
  \label{eq:qlearning}
\end{equation}
% where $\tilde{Q}(s, a)$ is the current estimate of the state-action value
% table.
The \textit{Bellman equation} \cite{sutton1998reinforcement} then takes the simple form
\begin{equation}
  \begin{split}
    Q^\pi(s, a) &= \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma \max_{a' \in \mathcal A}Q^\pi(s', a')]
    \\
    &= \mathbb E_{s' \sim p(s'|s,a)} [r(s,a) + \gamma Q^\pi(s', \pi(s'))],
  \end{split}
  \label{eq:bellman}
\end{equation}
a \textit{fixed-point} equation which provides a recursive decomposition of optimal behavior.
A complicated dynamic programming optimization can be
broken into simpler sub-problems at different time points,
which has often been closely related to the medial prefrontal part of the DMN
\citep{koechlin1999role, braver2002role}.
Using the Bellman equation, each state can be associated with a certain value,
to guide action towards a better state, thus improving on the agent's current policy.
% It is a fixed-point equation for the deterministic policy $\pi$.
Notice that in \eqref{eq:bellman}, the sampling is done over only things which
\textbf{XXX Ahhmm, please make a bigger effort here}
depend on the environment, and so can be learned off-policy by observing state-transitions
triggered by another behavioral policy, which can be stochastic.

\subsubsection{Efficient learning via value approximation and experience replay}
%As already mentioned in the previous section, Q-learning optimizes over the class of
% deterministic policies of the form \eqref{eq:qlearning}.
A successful RL theory needs ways to approximate the state-action value function
which scale up to large state and action spaces.
In state $s \in \mathcal S$, the agent takes an action $a \in \mathcal A$
% (for example, $a$ = ``Read'', or ``Run'', or ``Laugh'', or ``Sympathize'', or ``Empathize'', etc.)
by sampling from its current policy matrix, and collects a reward $r$,
% (measured on their internal currency of worldly values ?),
and the environment transitions to a new state $s' \in \mathcal S$. At the end of
this step, a new \textit{experience} $e = (s,a,r,s')$ is produced; this represents an exemplar
behavior of the agent and is recorded in replay memory buffer:
$D \leftarrow D \cup \{e\}$ (possibly discarding the oldest entries to make space).
Now, at iteration $k+1$, replay consists in sampling (uniform or importance-weighted
\footnote{e.g weighted by TD error of the state transition $s \overset{a}{\rightarrow} s'$.} ?)  mini-batches of experiences
$(s, a, r, s') \sim \mathcal U(D)$ from the replay memory $\mathcal D$
and trying to approximate
the would-be $Q$-value for the state-action pair $(s,a)$ as predicted by Bellman's equation \eqref{eq:bellman}, namely
$r + \gamma \max_{a'} Q(s', a'|\theta^Q_k)$, with the output of a parametrized regression model $(s,a,r,s')
\mapsto {Q}(s, a|\theta^Q_{k})$, where $Q(.,.|\theta^Q_{k})$ is an approximator for the $Q$-value operator,  parameterized by $\theta^Q_{k+1}$.
\textbf{XXX please cut previous sentence into smaller pieces}
